\documentclass{amsart}
\usepackage{appendix}

\usepackage[colorlinks=true,
citecolor=black,
linkcolor=black,
anchorcolor=black,
filecolor=black,
menucolor=black,
urlcolor=black,
pdftitle={Nilpotency, almost nonnegative curvature, and gradient flow},
pdfsubject={Differential geometry},
pdfauthor={Vitali Kapovitch, Anton Petrunin and Wilderich Tuschmann}
]{hyperref}

\begin{document}
\title
{On fundamental groups of towers of bundles}

\thanks{\it 2000 AMS Mathematics Subject Classification:\rm\
53C20. Keywords: nonnegative curvature, nilpotent}\rm
%\thanks{\it The first two authors were supported in part by NSF grants \# DMS-0204187 and \# DMS-0406482 respectively.The last author's research was partially supported by a DFG Heisenberg Fellowship.}
\author{Vitali Kapovitch}
\address{Vitali Kapovitch\\Department of Mathematics\\University of Toronto\\
Toronto, Ontario, M5S 2E4, Canada}\email{vtk@math.toronto.edu}
\author{Anton Petrunin }\address{Anton Petrunin\\ Department of Mathematics\\ Pennsylvania State University\\
University Park, State College, PA 16802
}\email{petrunin@math.psu.edu}
\author{Wilderich Tuschmann}\address{Wilderich Tuschmann
\\Mathematisches Seminar
\\Christian-Albrechts--Universit\"at zu Kiel\\ Ludewig-Meyn--Stra\ss{}e 4-8 \\ D-24118 Kiel,
Germany}\email{tusch@math.uni-kiel.de}

\maketitle
\begin{abstract}
We give a condition for towers of fiber bundles which implies that torsion of the fundamental group of its total space is nilpontent and its torsion lies in its center.
The theorem is motivated by a conjecture about almost nonnegatively curved manifolds.
\end{abstract}


\def\sectionthree{\cite[Section 3]{KPT} }
\def\kpt{ in \cite{KPT}}

\ 

\subsection{Introduction.}
This is an appendix to our paper on almost nonnegatively curved manifolds \cite{KPT}.
We prove the following topological theorem.
\input{appendix.tex}

\small
\bibliographystyle{alpha}
%\bibliography{kpt}

%\begin{comment}
\begin{thebibliography}{HMR75}
\bibitem[DDK]{DDK}{Dror, E. and Dwyer, W. G. and Kan, D. M.},
 \textit{Self-homotopy equivalences of virtually nilpotent spaces},
{Comment. Math. Helv.},
{56},
{1981},
{4},
pp. {599--614},

\bibitem[KPT]{KPT} Kapovitch, Vitali; Petrunin, Anton; Tuschmann, Wilderich Nilpotency, almost nonnegative curvature, and the gradient flow on Alexandrov spaces. Ann. of Math. (2) 171 (2010), no. 1, 343–373.
\end{thebibliography}

\end{document}