\documentclass{amsart}
\usepackage{appendix}
\usepackage[colorlinks=true,
citecolor=black,
linkcolor=black,
anchorcolor=black,
filecolor=black,
menucolor=black,
urlcolor=black,
pdftitle={On torsion in center conjecture},
pdfsubject={Differential geometry},
pdfauthor={Vitali Kapovitch, Anton Petrunin and Wilderich Tuschmann}
]{hyperref}

\begin{document}
\title[The torsion in the center conjecture]{On the\\ torsion in the center\\ conjecture} %\\
%(An appendix for ``Nilpotency, almost\\
%nonnegative curvature, and gradient flow'')}%
\thanks{\it 2000 AMS Mathematics Subject Classification:\rm\
53C20. Keywords: nonnegative curvature, nilpotent}\rm
\author{Vitali Kapovitch}
\address{Vitali Kapovitch\\Department of Mathematics\\University of Toronto\\
Toronto, Ontario, M5S 2E4, Canada.}\email{vtk@math.toronto.edu}
\author{Anton Petrunin }\address{Anton Petrunin\\ Department of Mathematics\\ Pennsylvania State University\\
University Park, State College, PA 16802, USA.
}\email{petrunin@math.psu.edu}
\author{Wilderich Tuschmann}\address{Wilderich Tuschmann
\\Arbeitsgruppe Differentialgeometrie
\\Institut f\"ur Algebra und Geometrie
\\Fakult\"at f\"ur Mathematik
\\Karlsruher Institut f\"ur Technologie
\\Englerstr. 2
\\D-76131 Karlsruhe, Deutschland.}\email{tuschmann@kit.edu}

\maketitle
\begin{abstract}
We present a condition for towers of fiber bundles which implies that the fundamental group of 
the total space has a nilpotent subgroup of finite index whose torsion is contained in its center.
Moreover, the index of the subgroup can be bounded in terms of the fibers of the tower.

Our result is motivated by the conjecture that
every almost nonnegatively curved closed $m$-dimensional manifold $M$ 
admits a finite cover $\tilde M$ for which the number of leafs is bounded in terms of $m$
such that the torsion of the fundamental group $\pi_1 \tilde M$ lies in its center.
\end{abstract}


\section{Introduction}

\noindent 
We prove the following topological result, which is motivated by Conjecture~\ref{con:tor} below.

\begin{thm}\label{thm:smooth}
Let $F_1,F_2,\dots,F_n$ be an array of closed manifolds 
such that each $F_i$ is either $\mathbb{S}^{1}$ or is simply connected. 
Assume $E$ is the total space of a tower of fiber bundles over a point
$$E=E_n\buildrel {F_n}\over \longrightarrow E_{n-1}\buildrel {F_{n-1}}\over\longrightarrow\dots\buildrel {F_1}\over\longrightarrow E_0=\{pt\}$$
and each of the bundles $E_k\buildrel {F_k}\over \longrightarrow E_{k-1}$ is homotopically trivial over the $1$-skeleton. 
Then the fundamental group $\pi_1E$ contains a nilpotent subgroup $\Gamma$ such that
$$[\pi_1E:\Gamma]\le \mathrm{Const}(F_1,F_2,\dots,F_n)\quad\text{and}\quad\mathrm{Tor}(\Gamma)\subset \mathrm{Z}(\Gamma),$$
where $\mathrm{Tor}(\Gamma)$ and $\mathrm{Z}(\Gamma)$ denote the torsion and the center of $\Gamma$, respectively.
%I changed $N$ to $\Gamma$ because later $N$ is used to denote a manifold.
\end{thm}

Our proof is surprisingly involved. 
We would be very much interested to see a proof based on a different idea;
in particular, it might help to establish the conjecture in  full generality.

\section{Motivation}

\subsection{} Theorem 1 is motivated by the following conjecture from \cite{KPT}.

\begin{mconj}\label{con:tor}
For all dimensions $m$ there is a constant $C=C(m)$ such that if $M^m$ is an almost nonnegatively curved 
closed smooth $m$-manifold, then there is a nilpotent subgroup $N\subset \pi_1M$ of index  at most $C$ whose torsion is contained in its center.
\end{mconj}

The following earlier conjecture of Fukaya and Yamaguchi \cite{FY} is closely related.

\begin{conj}[Fukaya--Yamaguchi]\label{con:c-ab}
The fundamental group of a nonnegatively curved  $m$-manifold $M$ is $C(m)-$abelian: 
there is $C=C(m)$ such that if $M^m$ is nonnegatively curved, then there is an 
abelian subgroup $A\subset \pi_1M$ of index  at most~$C$.
\end{conj}

Let us show that our main conjecture (if true) implies Conjecture \ref{con:c-ab}.

If the fundamental group of $M$ is finite (in particular if $M$ is positively curved), then whole group is torsion.
In this case Conjecture~\ref{con:tor} and Theorem B in \cite{KPT} imply Conjecture~\ref{con:c-ab}.
The general case is more tricky.
The following argument
was suggested to us by Burkhard Wilking (compare to \cite[Corollary 4.6.1]{KPT}).

If $\sec(M^m)\ge0$, then the universal cover $\tilde M$ of $M$ is isometric to the product $\mathbb{R}^n\times K$, where $K$ is a compact Riemannian manifold and the $\pi_1M$ action  on $\mathbb{R}^n\times K$ is diagonal~\cite{CG72}.
It follows from \cite[Cor. 6.3]{wilking} that one can deform the metric on $M$ so that its universal cover is still isometric to $\mathbb{R}^n\times K$ and the induced action on $K$ is finite.
Let $\Gamma_1$ be the image of $\pi_1M$ in $\Iso(\R^n)$ and $\Gamma_2$ be the image in $\Iso(K)$. By the above we can assume that $\Gamma_2$ is finite.
Therefore $\Gamma_1$ is a crystallographic group.
By the third Bieberbach theorem, it contains a subgroup of index $\le C(n)$ consisting entirely of translations (and hence is abelian).
Consider the action of $\Gamma_2$ on the frame bundle $F$ of $K$.
This action is free and by Cheeger's trick, $F$ admits a sequence of almost nonnegatively curved metrics~\cite{FY}. 
Since $\Gamma_2$ is finite, by Conjecture~\ref{con:tor} it contains 
an abelian subgroup of index $\le C(m)$. 
Thus, both $\Gamma_1$ and $\Gamma_2$ are $C(m)-$abelian,
and this yields Conjecture~\ref{con:c-ab}.

\subsection{} Let us explain  how Theorem~\ref{thm:smooth} is related to Conjecture~\ref{con:tor}.

An almost nonnegatively curved manifold can be defined as a closed smooth manifold $M$ which admits a sequence of metrics $g_n$ with a uniform lower curvature bound such that the sequence $M_n$ converges to a point in Gromov--Hausdorff topology.

One approach to studying almost nonnegatively curved manifolds is by studying successive blow-ups of the collapsing sequence $M_n=(M,g_n)$, as was done in \cite[Section 4.3]{KPT}.
Let us describe this construction.

The sequence $\{M_n\}$ converges to a point; this one-point space will be denoted by $A_0$.

Set $M_{n,1}:=M_n$ and $\vartheta_{n,1}:=\diam M_{n,1}$.
Rescale now $M_{n,1}$ by $\tfrac{1}{\vartheta_{n,1}}$
so that 
\[\diam(\tfrac{1}{\vartheta_{n,1}}\cdot M_{n,1})=1.\]
Passing to a subsequence if necessary, one has
that the manifolds $\frac{1}{\vartheta_{n,1}}{\cdot} M_{n,1}$
converge to $A_1$,
where $A_1$
is a compact nonnegatively curved Alexandrov space with diameter $1$.

Now choose a regular point $\bar p_1\in A_1$,
and consider distance coordinates around $\bar p_1\in U_1\to \mathbb{R}^{k_1}$,
where $k_1$ is the dimension of $A_1$.
The distance functions can be lifted
to $U_{n,1}\subset \frac{1}{\vartheta_{n,1}}{\cdot} M_{n,1}$.
Denote by $M_{n,2}$ the level set of $U_{n,1}\to \mathbb{R}^{k_1}$ that corresponds to $\bar p_1$.
Clearly, $M_{n,2}$ is a compact submanifold of codimension $k_1$.
Set $\vartheta_{n,2}:=\diam M_{n,2}$.
Passing again to a subsequence, one has
that the sequence $\frac{1}{\vartheta_{n,2}}{\cdot} M_{n,2}$
converges to some  Alexandrov space $A_2$.

As before, $A_2$ is a compact nonnegatively curved Alexandrov space with diameter 1.
Set $k_2:=k_1+\dim A_2$.
If one now chooses a marked point in $M_{n,2}$,
then, as $n\to\infty$,
$$\frac{1}{\vartheta_{n,2}}{\cdot} M_{n}$$ converges 
in the pointed Gromov--Hausdorff topology
to $A_2\times\mathbb{R}^{k_1}$,
which is of dimension $k_2$.

We repeat this procedure until, at some step, $k_\ell=m$ where $m=\dim M$.

\medskip

As a result, we obtain a sequence $\{A_i\}$ of compact nonnegatively curved Alexandrov spaces
with diameter $1$ that satisfies
$$\dim A_i=k_i-k_{i-1} 
\quad\text{ and therefore }\quad
\sum_{i=1}^\ell\dim A_i=m.$$
We also obtain
a sequence of rescaling factors $\vartheta_{n,i}=\diam M_{n,i}$,
and  a nested sequence of submanifolds
$$\{p_n\}=M_{n,\ell}\subset \cdots \subset M_{n,2}\subset M_{n,1}=M_n,$$

Now let us assume that \emph{for all choices of $\bar p_i$}
the Alexandrov spaces $A_1,\dots, A_\ell$ are closed  Riemannian manifolds.
This condition will be called the \emph{rescaled smoothness assumption}.

Notice tbat his assumption definitely \emph{does not need to hold} in general;
it is only made to simplify the problem.
Foremost, it makes it possible to apply the following fibration theorem of Yamaguchi~\cite{Yam} to the successive blow-ups of the collapsing sequence under consideration.

\begin{thm}[Yamaguchi's Fibration Theorem]
Let $M_n$ be a sequence of $m$-dimensional Riemannian manifolds with sectional curvatue at least $k$ 
and diameters at most $D$, which converges to a Riemannian manifold $N$ in the Gromov--Hausdorff sense.
Then for all large $n$ there exists an almost Riemannian submersion $f_n\co M_n\to N$.
In particular, $f_n$ is a fiber bundle.
\end{thm}

Let $M_n$ as above satisfies the rescaled smoothness assumption.
Then it should be true that after passing to a subsequence of $M_n$,
each $M_n$ is homeomorphic to the total space of a tower of fiber bundles
\[
M_n=E_\ell\buildrel {A_\ell}\over \longrightarrow E_{n-1}\buildrel {A_{\ell-1}}\over\longrightarrow\dots\buildrel {A_1}\over\longrightarrow E_0=\{pt\}.
\]
Moreover, by using the Lipschitz properties of the gradient flow,
it should be possible to pass to a finite cover of the total space with number of leafs 
controlled by the topology of $A_i$, so that the tower of bundles can be
further refined  to a tower of the form
\[
E=E_n\buildrel {F_n}\over \longrightarrow E_{n-1}\buildrel {F_{n-1}}\over\longrightarrow\dots\buildrel {F_1}\over\longrightarrow E_0=\{pt\}
\]
such that
\begin{enumerate}[(i)]
\item Each $F_i$ is either $\mathbb{S}^{1}$ or is simply connected. 
\item Each of the bundles $E_k\buildrel {F_k}\over \longrightarrow E_{k-1}$ is homotopically trivial over the $1$-skeleton. 
\end{enumerate}

This finally leads us to the main topological result of this paper (Theorem \ref{thm:smooth} above),
which says that if such a tower of bundles exists, then its total space satisfies the conclusion of Conjecture~\ref{con:tor}.

We have, however, not worked out the details of the reduction, 
as we do not see a geometric condition which would imply the rescaled smoothness assumption.
In fact, it might  well be possible to do the reduction even in the case if all $A_i$ are general  Alexandrov spaces  \emph{without  boundary}, but we have no idea what to do if at least one of them has nonempty boundary.


\section{Proof of the theorem}

\subsection{}
The proof of Theorem~\ref{thm:smooth} uses the following result of Dror,  Dwyer and Kan (\cite{DDK}).

\input{appendix.tex}

\small
\bibliographystyle{alpha}
%\bibliography{master-july-2017}


\begin{thebibliography}{HMR75}

\bibitem[CG]{CG72}Cheeger, J.; Gromoll, D.
 \textit{On the structure of complete manifolds of nonnegative curvature.}
Ann. of Math. (2) 96 (1972), 413--443

\bibitem[DDK]{DDK}{Dror, E. and Dwyer, W. G. and Kan, D. M.},
 \textit{Self-homotopy equivalences of virtually nilpotent spaces},
{Comment. Math. Helv.},
{56},
{1981},
{4},
pp. {599--614},

\bibitem[FY]{FY}
K.~Fukaya and T.~Yamaguchi.
\newblock The fundamental groups of almost nonnegatively curved manifolds.
\newblock {\em Ann. of Math. (2)}, 136(2):253--333, 1992.


\bibitem[KPT]{KPT} Kapovitch, Vitali, Anton Petrunin, and Wilderich Tuschmann. 
\newblock Nilpotency, almost nonnegative curvature, and the gradient flow on Alexandrov spaces. 
\newblock Annals of Mathematics (2010), 343--373.

\bibitem[W]{wilking}
Wilking, Burkhard. \newblock On fundamental groups of manifolds of nonnegative curvature. \newblock  Differential Geom. Appl.  13  (2000),  no. 2, 129--165.

\bibitem[Y]{Yam}
T.~Yamaguchi.
\newblock Collapsing and pinching under a lower curvature bound.
\newblock {\em Ann. of Math.}, 133:317--357, 1991.
\end{thebibliography}


\end{document}